---
title: "oak_project"
output: html_document
date: "2024-05-20"
---

Genome explorations of Oak trees in Hainault forest National Park. 

# 1 PCA analyses

```{r, include=FALSE}

setwd(".")

# 1 Separating Genotype Matrix by Species

# Read the sample metadata
sample_metadata <- read.csv("samples-sheet.csv", header = TRUE, stringsAsFactors = FALSE)

## HERE FOR DIFFERENT CHROMOSOME

# Read the genotype matrix
genotype_matrix <- read.table("genotype-matrix-Qrob_Chr11", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE, sep="\t")

# Inspect the data
print("Sample Metadata:")
print(head(sample_metadata))

print("Genotype Matrix:")
print(head(genotype_matrix))

# Ensure the sample ID column in sample metadata is named "Sample_ID"
names(sample_metadata)[1] <- "Sample_ID"

# Create a named vector of species for easy lookup
sample_species <- setNames(sample_metadata$Species, sample_metadata$Sample_ID)

# Separate columns by species
quercus_robur_cols <- names(genotype_matrix) %in% names(sample_species[sample_species == "Quercus robur"])
quercus_petraea_cols <- names(genotype_matrix) %in% names(sample_species[sample_species == "Quercus petraea"])

# Create separate data frames for each species
quercus_robur_genotype <- genotype_matrix[, quercus_robur_cols]
quercus_petraea_genotype <- genotype_matrix[, quercus_petraea_cols]

# Inspect the separated data
print("Quercus robur Genotype Matrix:")
print(head(quercus_robur_genotype))

print("Quercus petraea Genotype Matrix:")
print(head(quercus_petraea_genotype))

# Save the separated genotype matrices
write.csv(quercus_robur_genotype, "quercus_robur_genotype_matrix.csv", row.names = FALSE)
write.csv(quercus_petraea_genotype, "quercus_petraea_genotype_matrix.csv", row.names = FALSE)

```

# 2.1 F_ST Scores Analysis

``` {r, echo=FALSE}

# 1 Preparing data for FST study

# Dependency: 1.1

# Load necessary libraries
library(dplyr)
library(ggplot2)

# Load genotype_matrix and sample_metadata
# Assuming they are already loaded into the environment

# Check the structures
head(genotype_matrix)
head(sample_metadata)

# Verify the subsets
head(quercus_robur_genotype)
head(quercus_petraea_genotype)

# Changing the name of the objects just cuz
genotype_robur <- quercus_robur_genotype
genotype_petraea <- quercus_petraea_genotype

```

## 2.2 Combining the datasets, adding labels and transposing

``` {r, echo=FALSE}

# 2.2 Combining the datasets

# Transpose genotype matrices to have samples as rows
genotype_robur_t <- as.data.frame(t(genotype_robur))
genotype_petraea_t <- as.data.frame(t(genotype_petraea))

# Ensure all genotype columns are numeric
genotype_robur_t[] <- lapply(genotype_robur_t, as.numeric)
genotype_petraea_t[] <- lapply(genotype_petraea_t, as.numeric)

# Add Population column
genotype_robur_t$Population <- "Quercus robur"
genotype_petraea_t$Population <- "Quercus petraea"

# Combine the datasets
combined_genotype <- rbind(genotype_robur_t, genotype_petraea_t)

# Ensure Population column is a factor
combined_genotype$Population <- as.factor(combined_genotype$Population)

```
``` {r, echo=FALSE}

m <- combined_genotype

# Check if all sample IDs in the genotype matrix match the sample metadata
matching_ids <- rownames(m) %in% sample_metadata$Sample_ID
if (!all(matching_ids)) {
  stop("Some sample IDs in the genotype matrix do not match the sample metadata.")
}

# Subset the sample metadata to match the genotype matrix rows
sample_metadata <- sample_metadata[match(rownames(m), sample_metadata$Sample_ID), ]

# Ensure that the number of rows matches between the matrix and metadata
if (nrow(m) != nrow(sample_metadata)) {
  stop("Number of rows in the genotype matrix does not match the sample metadata after subsetting.")
}


# Convert all columns to numeric
m_numeric <- apply(m, 2, function(x) as.numeric(as.character(x)))

# Check for non-numeric columns that were converted to NA
non_numeric_cols <- apply(m_numeric, 2, function(x) any(is.na(x)))
if (any(non_numeric_cols)) {
  warning("There are non-numeric columns that were converted to NA. These will be removed.")
}

# Remove columns that became NA due to non-numeric conversion
m_numeric <- m_numeric[, !non_numeric_cols]

# Check the dimensions of the numeric matrix
dim(m_numeric)

# Custom scaling function
scale_custom <- function(x) {
  if (sd(x, na.rm = TRUE) == 0) {
    return(x)  # Return as is if variance is zero
  } else {
    return((x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))  # Standard scale
  }
}

# Apply custom scaling function to the numeric matrix
m_scaled <- apply(m_numeric, 2, scale_custom)

# Check the dimensions of the scaled matrix
dim(m_scaled)

# Perform PCA on the scaled matrix
p <- prcomp(m_scaled, retx = TRUE)

# Extract variance explained by each principal component
v <- p$sdev^2
pv <- 100 * v / sum(v)

# Plot the percentage of variance explained by each principal component
barplot(pv, xlab = "PC", ylab = "% variance", ylim = c(0, 100), main = "% Variance explained by each principal component (PC)")

# Combine PCA scores with sample metadata
df_PCA <- cbind(sample_metadata, p$x[, 1:4]) # Adjust number of PCs as needed

# Check the dimensions of the combined data frame
dim(df_PCA)

# Plot the PCA results using ggplot2
library(ggplot2)
ggplot(df_PCA, aes(x = PC1, y = PC2, color = Species)) + 
  geom_point() + 
  theme_minimal() + 
  labs(title = "PCA of Genotype Data", x = paste0("PC1 (", round(pv[1], 1), "%)"), y = paste0("PC2 (", round(pv[2], 1), "%)")) +
  theme(legend.position = "right")


```

Most data points for each species are concentrated at very low values on both axes, with just one extreme outlier on the PC2 axis.

This outlier (a point far along PC2) could be an anomaly, possibly representing a different population or an issue in the data.

Typically, hybridization would be suggested by intermediate positions between clusters, indicating genetic overlap or blending. However, the points here are either highly clustered near the origin or isolated, with no gradual distribution.

ntrogression, or the incorporation of genetic material from one species into another through backcrossing, might show some intermediate positioning of points across both axes. However, this plot doesn’t offer clear evidence of this.

The tight clustering near the origin and the isolated outlier suggest possible issues with data scaling or outlier handling, rather than direct evidence of genetic mixing.


``` {r, echo=FALSE, eval=FALSE}

# Load necessary library
data <- read.table("genotype-matrix-single-file", header = TRUE, stringsAsFactors = FALSE, check.names = FALSE, sep="\t")

# Extract genotype columns (assuming first four columns are CHROM, POS, REF, ALT)
genotype_data <- data[, 5:ncol(data)]

# Convert genotype data to numeric
genotype_data <- as.data.frame(lapply(genotype_data, function(x) as.numeric(as.character(x))))

# Check the structure of the genotype data
str(genotype_data)

# Transpose the genotype data
genotype_matrix <- t(genotype_data)

# Assign row names (sample IDs) from the original column names
rownames(genotype_matrix) <- colnames(data)[5:ncol(data)]

# Convert to data frame for PCA
genotype_matrix <- as.data.frame(genotype_matrix)

# Check the dimensions of the genotype matrix
dim(genotype_matrix)

# Custom scaling function to handle zero variance columns
scale_custom <- function(x) {
  if (sd(x, na.rm = TRUE) == 0) {
    return(x)  # Return as is if variance is zero
  } else {
    return((x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))  # Standard scale
  }
}

# Apply custom scaling function to the genotype matrix
genotype_scaled <- as.data.frame(apply(genotype_matrix, 2, scale_custom))

# Perform PCA on the scaled genotype matrix
pca_result <- prcomp(genotype_scaled, retx = TRUE)

# Extract variance explained by each principal component
variance_explained <- pca_result$sdev^2
percent_variance <- 100 * variance_explained / sum(variance_explained)

# Plot the percentage of variance explained by each principal component
barplot(percent_variance, xlab = "PC", ylab = "% Variance", ylim = c(0, 100), main = "% Variance explained by each principal component (PC)")

# Combine PCA scores with sample metadata if available
# Assuming sample_metadata is available and matches the rownames of genotype_matrix
# df_PCA <- cbind(sample_metadata, pca_result$x[, 1:4]) # Adjust number of PCs as needed

# Plot the PCA results using ggplot2
library(ggplot2)
ggplot(data.frame(pca_result$x), aes(x = PC1, y = PC2)) + 
  geom_point() + 
  theme_minimal() + 
  labs(title = "PCA of Genotype Data", x = paste0("PC1 (", round(percent_variance[1], 1), "%)"), y = paste0("PC2 (", round(percent_variance[2], 1), "%)")) +
  theme(legend.position = "right")

```

``` {r, echo=FALSE,eval=FALSE}

# The code below processes the genotype matrix and performs PCA on the data. 
# Not working due to there being too many duplicate sample IDs in the metadata, so it was discarded.

#

# Assuming sample_metadata is your metadata dataframe
# and it contains a column named "Sample_ID" and "Species"

# Ensure the row names of the genotype matrix are set to match sample IDs
rownames(genotype_matrix) <- colnames(data)[5:ncol(data)]

# Align the sample metadata with the genotype matrix rows
sample_metadata <- sample_metadata[match(rownames(genotype_matrix), sample_metadata$Sample_ID), ]

# Ensure that the number of rows matches between the matrix and metadata
if (nrow(genotype_matrix) != nrow(sample_metadata)) {
  stop("Number of rows in the genotype matrix does not match the sample metadata.")
}

# Convert all columns to numeric
genotype_numeric <- apply(genotype_matrix, 2, function(x) as.numeric(as.character(x)))

# Check for non-numeric columns that were converted to NA
non_numeric_cols <- apply(genotype_numeric, 2, function(x) any(is.na(x)))
if (any(non_numeric_cols)) {
  warning("There are non-numeric columns that were converted to NA. These will be removed.")
}

# Remove columns that became NA due to non-numeric conversion
genotype_numeric <- genotype_numeric[, !non_numeric_cols]

# Check the dimensions of the numeric matrix
dim(genotype_numeric)

# Custom scaling function
scale_custom <- function(x) {
  if (sd(x, na.rm = TRUE) == 0) {
    return(x)  # Return as is if variance is zero
  } else {
    return((x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))  # Standard scale
  }
}

# Apply custom scaling function to the numeric matrix
genotype_scaled <- as.data.frame(apply(genotype_numeric, 2, scale_custom))

# Check the dimensions of the scaled matrix
dim(genotype_scaled)

# Perform PCA on the scaled matrix
pca_result <- prcomp(genotype_scaled, retx = TRUE)

# Extract variance explained by each principal component
variance_explained <- pca_result$sdev^2
percent_variance <- 100 * variance_explained / sum(variance_explained)

# Plot the percentage of variance explained by each principal component
barplot(percent_variance, xlab = "PC", ylab = "% Variance", ylim = c(0, 100), main = "% Variance explained by each principal component (PC)")

# Combine PCA scores with sample metadata
df_PCA <- cbind(sample_metadata, pca_result$x[, 1:4]) # Adjust number of PCs as needed

# Check the dimensions of the combined data frame
dim(df_PCA)

# Plot the PCA results using ggplot2
library(ggplot2)
ggplot(df_PCA, aes(x = PC1, y = PC2, color = Species)) + 
  geom_point(size = 2) + 
  theme_minimal() + 
  labs(title = "PCA of Genotype Data - All Chromosomes", x = paste0("PC1 (", round(percent_variance[1], 1), "%)"), y = paste0("PC2 (", round(percent_variance[2], 1), "%)")) +
  theme(legend.position = "right")

###

# Check for duplicate Sample_IDs in sample_metadata
duplicate_ids <- sample_metadata$Sample_ID[duplicated(sample_metadata$Sample_ID)]
if (length(duplicate_ids) > 0) {
  stop(paste("Duplicate Sample_IDs found:", paste(duplicate_ids, collapse = ", ")))
}

# Assuming sample_metadata is your metadata dataframe
# and it contains a column named "Sample_ID" and "Species"

# Ensure the row names of the genotype matrix are set to match unique sample IDs
rownames(genotype_matrix) <- sample_metadata$Sample_ID

# Align the sample metadata with the genotype matrix rows
sample_metadata <- sample_metadata[match(rownames(genotype_matrix), sample_metadata$Sample_ID), ]

# Ensure that the number of rows matches between the matrix and metadata
if (nrow(genotype_matrix) != nrow(sample_metadata)) {
  stop("Number of rows in the genotype matrix does not match the sample metadata.")
}

# Convert all columns to numeric
genotype_numeric <- apply(genotype_matrix, 2, function(x) as.numeric(as.character(x)))

# Check for non-numeric columns that were converted to NA
non_numeric_cols <- apply(genotype_numeric, 2, function(x) any(is.na(x)))
if (any(non_numeric_cols)) {
  warning("There are non-numeric columns that were converted to NA. These will be removed.")
}

# Remove columns that became NA due to non-numeric conversion
genotype_numeric <- genotype_numeric[, !non_numeric_cols]

# Check the dimensions of the numeric matrix
dim(genotype_numeric)

# Custom scaling function
scale_custom <- function(x) {
  if (sd(x, na.rm = TRUE) == 0) {
    return(x)  # Return as is if variance is zero
  } else {
    return((x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE))  # Standard scale
  }
}

# Apply custom scaling function to the numeric matrix
genotype_scaled <- as.data.frame(apply(genotype_numeric, 2, scale_custom))

# Check the dimensions of the scaled matrix
dim(genotype_scaled)

# Perform PCA on the scaled matrix
pca_result <- prcomp(genotype_scaled, retx = TRUE)

# Extract variance explained by each principal component
variance_explained <- pca_result$sdev^2
percent_variance <- 100 * variance_explained / sum(variance_explained)

# Plot the percentage of variance explained by each principal component
barplot(percent_variance, xlab = "PC", ylab = "% Variance", ylim = c(0, 100), main = "% Variance explained by each principal component (PC)")

# Combine PCA scores with sample metadata
df_PCA <- cbind(sample_metadata, pca_result$x[, 1:4]) # Adjust number of PCs as needed

# Check the dimensions of the combined data frame
dim(df_PCA)

# Plot the PCA results using ggplot2
library(ggplot2)
pca_plot <- ggplot(df_PCA, aes(x = PC1, y = PC2, color = Species)) + 
  geom_point(size = 2) + 
  theme_minimal() + 
  labs(title = "PCA of Genotype Data - All Chromosomes", x = paste0("PC1 (", round(percent_variance[1], 1), "%)"), y = paste0("PC2 (", round(percent_variance[2], 1), "%)")) +
  theme(legend.position = "right")
print(pca_plot)

# Manually identify outliers based on visual inspection of the plot
# Suppose you identified samples with IDs 10, 20, 30 as outliers
outlier_ids <- c(10, 20, 30)

# Ensure sample IDs are character for matching
outlier_ids <- as.character(outlier_ids)

# Remove outliers from genotype data and metadata using Sample_IDs
genotype_cleaned <- genotype_numeric[!rownames(genotype_numeric) %in% outlier_ids, ]
sample_metadata_cleaned <- sample_metadata[!sample_metadata$Sample_ID %in% outlier_ids, ]

# Ensure the dimensions match after removing outliers
if (nrow(genotype_cleaned) != nrow(sample_metadata_cleaned)) {
  stop("Number of rows in the cleaned genotype matrix does not match the cleaned sample metadata.")
}

# Standardize the cleaned data
genotype_cleaned_scaled <- as.data.frame(apply(genotype_cleaned, 2, scale_custom))

# Recompute PCA on cleaned data
pca_cleaned_result <- prcomp(genotype_cleaned_scaled, retx = TRUE)

# Extract variance explained by each principal component for the cleaned data
variance_explained_cleaned <- pca_cleaned_result$sdev^2
percent_variance_cleaned <- 100 * variance_explained_cleaned / sum(variance_explained_cleaned)

# Plot the percentage of variance explained by each principal component for the cleaned data
barplot(percent_variance_cleaned, xlab = "PC", ylab = "% Variance", ylim = c(0, 100), main = "% Variance explained by each principal component (PC) - Cleaned")

# Combine PCA scores with cleaned sample metadata
df_PCA_cleaned <- cbind(sample_metadata_cleaned, pca_cleaned_result$x[, 1:4]) # Adjust number of PCs as needed

# Plot the cleaned PCA results using ggplot2
pca_cleaned_plot <- ggplot(df_PCA_cleaned, aes(x = PC1, y = PC2, color = Species)) + 
  geom_point(size = 2) + 
  theme_minimal() + 
  labs(title = "PCA of Genotype Data - All Chromosomes (Cleaned)", x = paste0("PC1 (", round(percent_variance_cleaned[1], 1), "%)"), y = paste0("PC2 (", round(percent_variance_cleaned[2], 1), "%)")) +
  theme(legend.position = "right")
print(pca_cleaned_plot)

```

## 2.3 Calculating F_ST using heirfstat package, 

Calculates the overall FST value on the entire chromosome using the hierfstat package. This workflow helps assess the degree of genetic differentiation between populations across the chromosome based on the FST metric. This is a time consuming process.

``` {r, echo=FALSE}

# 2.3 Calculating overall F_ST for the whole chromosome

# Install and load the hierfstat package if not already installed
if (!requireNamespace("hierfstat", quietly = TRUE)) {
  install.packages("hierfstat")
}
library(hierfstat)

# Prepare the data for hierfstat
# Ensure Population is the first column and all other columns are numeric
hierfstat_data <- combined_genotype %>% 
  dplyr::select(Population, everything())

# Verify the structure of hierfstat_data, ensuring that it is set up correctly
str(hierfstat_data)

# Calculate F_ST
fst_results <- wc(hierfstat_data, diploid = TRUE)
fst_value <- fst_results$FST
head(fst_value)

```

Chloroplast FST is `0.0003919184`
Mitochondria FST is `0.001702633`
Chromosome 10 FST is `0.005184404`
Chromosome 11 FST is `0.002538074`
The low FST indicates a very low level of genetic differentiation between sessile and pedunculate oak populations.

This result was not published in our final report as FST is not a sensitive enough metric to be accurate in determining the level of genetic differentiation between the two oak species, likely due to significant genetic overlap between the two species (best demonstrated in our figures).

But it adds additional statistical rigor to support our results.

``` {r, include=FALSE, eval=FALSE}

# NOT WORKING

# 2.4 Calculating FST for each SNP loci separately

# Function to calculate F_ST for a single SNP
calculate_fst_snp <- function(genotype_data, population) {
  snp_data <- data.frame(genotype = as.numeric(genotype_data), population = population)
  snp_data <- snp_data[complete.cases(snp_data), ] # Remove rows with NAs
  fst_result <- wc(snp_data, diploid = TRUE)
  return(fst_result$FST)
}

# Extract population data
population_data <- hierfstat_data$Population

# Calculate F_ST for each SNP
fst_values <- apply(hierfstat_data[, -1], 2, calculate_fst_snp, population = population_data)

# Verify F_ST values
head(fst_values)

```

## 2.5 Watterson's Theta and Charts

``` {r, echo=FALSE}

# 2.5 Producing a data frame of Watterson's Theta values (NOT WORKING)

# Function to calculate Watterson's Theta
calculate_theta <- function(genotype_matrix) {
  S <- apply(genotype_matrix, 2, function(row) sum(row != 0 & row != 2)) # Segregating sites
  n <- nrow(genotype_matrix) # Number of sequences
  a_n <- sum(1 / (1:(n - 1))) # Harmonic number
  theta <- S / a_n
  return(theta)
}

# Calculate Theta for each species
theta_robur <- calculate_theta(genotype_robur_t[, -ncol(genotype_robur_t)])
theta_petraea <- calculate_theta(genotype_petraea_t[, -ncol(genotype_petraea_t)])

head(theta_robur)
head(theta_petraea)
```
``` {r, echo=FALSE}
# Charts for Watterson's Theta values

# 1 barplot (Small datasets)

# Combine the Theta values into a data frame
theta_data <- data.frame(
  SNP = 1:length(theta_robur),
  Theta_Robur = theta_robur,
  Theta_Petraea = theta_petraea
)

# Melt the data frame for ggplot2
library(reshape2)
theta_melted <- melt(theta_data, id.vars = "SNP", variable.name = "Species", value.name = "Theta")

# Create a bar plot
library(ggplot2)
ggplot(theta_melted, aes(x = factor(SNP), y = Theta, fill = Species)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  xlab("SNP") +
  ylab("Watterson's Theta") +
  ggtitle("Watterson's Theta Values for Quercus robur and Quercus petraea") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(values = c("Theta_Robur" = "blue", "Theta_Petraea" = "red"))
```

Quercus robur (in blue) generally shows higher Watterson's Theta values than Quercus petraea (in red) at many SNP sites. This could suggest that Q. robur has higher genetic diversity in this dataset, although you'd need to calculate average values for each species to confirm this systematically.

The variation in Watterson's Theta across SNP sites indicates that genetic diversity is not uniform across the genome. Some SNPs have much higher Theta values, which might indicate areas with higher polymorphism rates or regions with historical selective pressures.

The presence of a few high peaks (where Theta values reach above 8 or even 12) could point to regions with a high concentration of genetic variation in Q. robur. This could potentially be regions under balancing selection or areas where recombination is frequent.

Additional statistical analyses, such as mean and variance comparisons, are needed.

``` {r, echo=FALSE, eval=FALSE}

# 2 boxplot (Small datasets)

# Create a box plot
ggplot(theta_melted, aes(x = Species, y = Theta, fill = Species)) +
  geom_boxplot() +
  theme_minimal() +
  xlab("Species") +
  ylab("Watterson's Theta") +
  ggtitle("Distribution of Watterson's Theta Values") +
  scale_fill_manual(values = c("Theta_Robur" = "blue", "Theta_Petraea" = "red"))

```

``` {r, echo=FALSE, eval=FALSE}

# 3 Aesthetically appealing (not really) boxplot (Small datasets)

# Load necessary libraries
library(dplyr)
library(reshape2)
library(ggplot2)

# Combine the Theta values into a data frame
theta_data <- data.frame(
  SNP = 1:length(theta_robur),
  Theta_Robur = theta_robur,
  Theta_Petraea = theta_petraea
)

# Melt the data frame for ggplot2
theta_melted <- melt(theta_data, id.vars = "SNP", variable.name = "Species", value.name = "Theta")

# Calculate counts of data points for each species
counts <- theta_melted %>%
  group_by(Species) %>%
  summarise(count = n())

# Create an improved box plot with jittered points and count annotations
ggplot(theta_melted, aes(x = Species, y = Theta, fill = Species)) +
  geom_boxplot(outlier.shape = NA) + # Remove outliers from the box plot
  geom_jitter(width = 0.2, alpha = 0.5) + # Add jittered points
  theme_minimal() +
  xlab("Species") +
  ylab("Watterson's Theta") +
  ggtitle("Distribution of Watterson's Theta Values") +
  scale_fill_manual(values = c("Theta_Robur" = "blue", "Theta_Petraea" = "red")) +
  ylim(0, 7.5) + # Set y-axis limit
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Adjust x-axis text
  geom_text(data = counts, aes(x = Species, y = 7.5, label = paste0("n=", count)), vjust = -0.5) # Add counts

```